Bayesian Sparsification of $\cplx$-valued networks

### 2019-11-04 paper/paper.tex
rewrote sections on CR-calculus; moved cplx criticism to discussion

% complex nets and related stuff
Maths and related results \cite{pav_moments_2015,taubock_complex-valued_2012}, and
\cite{karseras_caution:_2014}

@@

The question of differentiability of complex-valued networks is tricky, since they usually
use non-differentiable non-linearities. However, a more severe problem stems from the loss
objective being real-valued. Indeed, if $f\colon \cplx \to \cplx$ were $\cplx$-differentiable
(holomorphic), the composition of $f$ with a non-trivial $h \colon \cplx\to \real$ makes it
not $\cplx$-differentiable with respect parameters, since $h$ violates Cauchy-Riemann conditions.
This problem is commonly dealt with by using Wirtinger, or $\cplx\real$ calculus, which subsumes
$\cplx$-calculus, but is also applicable to non-holomorphic functions of $\cplx$-argument, see
\cite{adali_complex-valued_2011} and \cite{trabelsi_deep_2017}. In essence, this approach
regards $f$ as a function of two independent variables $z$ and its conjugate $\conj{z}$ and
symbolically redefines the derivatives appropriately. The resulting derivative operators
satisfy linearity, product and chain rules, and respect complex conjugation. This approach
can be directly reformulated in terms of derivatives with respect to real and imaginary parts,
making it possible to use and apply the rules of $\real$-calculus to the networks represented
in paired $\real$ fashion and use existing $\real$-valued backpropagation in autograd software.

@@

@@

TLDR: we use Wirtinger calculus!

% essential intro into Wirtinger calculus (CR)
Take $f\colon \cplx \to \cplx$ and identify it with $F\colon \real^2 \to \cplx$ function
via $F(u, v) = f(z)\vert_{z=u + j v}$. Wirtinger calculus introduces two formal derivative
operators \rewrite{(this seems to be very verbatim)}
$
  \tfrac{\partial}{\partial z}
    = \tfrac12 \bigl(
      \tfrac{\partial}{\partial u}
      - j \tfrac{\partial}{\partial v}
    \bigr)
$ and $
  \tfrac{\partial}{\partial \conj{z}}
    = \tfrac12 \bigl(
      \tfrac{\partial}{\partial u}
      + j \tfrac{\partial}{\partial v}
    \bigr)
$, defines $dz = du + j dv$, $d\conj{z} = du - j dv$, and the differential of $f$ as
$$
df = \tfrac{\partial f}{\partial z} dz
    + \tfrac{\partial f}{\partial \conj{z}} d\conj{z}
   % = \tfrac12\bigl(
   %    \tfrac{\partial}{\partial u}
   %    - j \tfrac{\partial}{\partial v}
   % \bigr) F (du - j dv)
   % + \tfrac12\bigl(
   %    \tfrac{\partial}{\partial u}
   %    + j \tfrac{\partial}{\partial v}
   % \bigr) F (du - j dv)
   % = \tfrac12 \bigl(
   %    \tfrac{\partial F}{\partial u} du - j \tfrac{\partial F}{\partial v} du
   %    + j \tfrac{\partial F}{\partial u} dv + \tfrac{\partial F}{\partial v} dv
   % \bigr)
   % + \tfrac12 \bigl(
   %    \tfrac{\partial F}{\partial u} du + j \tfrac{\partial F}{\partial v} du
   %    - j \tfrac{\partial F}{\partial u} dv + \tfrac{\partial F}{\partial v} dv
   % \bigr)
   = \tfrac{\partial F}{\partial u} du
     + \tfrac{\partial F}{\partial v} dv
   = dF
  \,. $$
Thus the complex value and its conjugate are treated as independent variables. In this notation
Cauchy-Riemann conditions are expressed as $
  \tfrac{\partial f}{\partial \conj{z}} = 0
$, or $
  -j\tfrac{\partial F}{\partial v} = \tfrac{\partial F}{\partial v}
$. These conditions impose a rigid structure on real and imaginary parts of $F(u, v)$, namely $
  \tfrac{\partial F_{\Re }}{\partial u} = \tfrac{\partial F_{\Im }}{\partial v}
$ and $
  \tfrac{\partial F_{\Re }}{\partial v} = - \tfrac{\partial F_{\Im }}{\partial u}
$. Thus Wirtinger calculus subsumes the usual $\cplx$-calculus, when $
  \tfrac{\partial f}{\partial \conj{z}} = 0
$, i.e. $
  f(z)
    % = g(z, \conj{z})
    = F(\tfrac12 (z + \conj{z}), \tfrac1{j 2} (z - \conj{z}))
$ does not depend on $\conj{z}$.

For a real-valued $f\colon \cplx \to \real$ we have $\conj{f} = f$, whence $
  \tfrac{\partial f}{\partial \conj{z}}
    = \tfrac{\partial \conj{f}}{\partial \conj{z}}
    = \conj{\tfrac{\partial f}{\partial z}}
$. Hence
$$
df
  = \tfrac{\partial f}{\partial z} dz
    + \tfrac{\partial f}{\partial \conj{z}} d\conj{z}
  % = \conj{\tfrac{\partial \conj{f}}{\partial \conj{z}}} dz
  % + \conj{\conj{\tfrac{\partial f}{\partial \conj{z}}} dz}
  = 2 \Re \Bigl(
    \conj{\tfrac{\partial f}{\partial \conj{z}}} dz
  \Bigr)
  = 2 \Re \bigl(
    \tfrac{\partial f}{\partial z} dz
  \bigr)
  \,. $$
Hence the gradient, being the direction of \rewrite{maximal growth} of $f$ at $z$ is given by $
  \tfrac{\partial f}{\partial \conj{z}}
    = \tfrac{\partial F}{\partial u}
      + j \tfrac{\partial F}{\partial v}
$. Thus under Wirtinger calculus it is possible to reuse $\real$-backpropagation.

% Show that $dh(z) = df(g(z)) dg(z)$.
% $h(z) = f(g(z)) = F(G_r(u, v), G_i(u, v)) = H(u, v)$
% $$
% dH
%   = \tfrac{\partial F}{\partial x} \tfrac{\partial G_r}{\partial u} du
%   + \tfrac{\partial F}{\partial x} \tfrac{\partial G_r}{\partial v} dv
%   + \tfrac{\partial F}{\partial y} \tfrac{\partial G_i}{\partial u} du
%   + \tfrac{\partial F}{\partial y} \tfrac{\partial G_i}{\partial v} dv
%   = \tfrac{\partial F}{\partial r} d\Re g(z)
%   + \tfrac{\partial F}{\partial i} d\Im g(z)
%   = df(\Re g(z) + j\Im g(z)) d g(z)
%   = df(g(z)) d g(z)
%   \,. $$

>>>>

% complex nets and related stuff
data that is naturally complex-valued
studies have shown a potentially richer representation of complex nets
\cite{hirose2003,fijulamin2009}
% [3] Akira Hirose, Complex-valued neural networks: theories and applications, vol. 5, World Scientific, 2003.
% [4] Md Faijul Amin and Kazuyuki Murase, “Single-layered complex-valued neural network for real-valued classification problems,” Neurocomputing, vol. 72, no. 4-6, pp. 945–955, 2009.

the complex version of backpropagation algorithm  was proposed because of
its richer representational capacity.
% Nevio Benvenuto and Francesco Piazza, “On the complex backpropagation algorithm,” IEEE Transactions on Signal Processing, vol. 40, no. 4, pp. 967–969, 1992.

\cite{yang_complex_2019} develop complex transformer, with $\cplx$-attention and complex
encoder-decoder, and apply it to music transcription and wireless signal classification.

that interfaces the rest of
$\cplx$-valued
the network with complex numbers

that operates on the Fourier transform input speech, acoustic, of radio signals.

papers related to complex-valued distributions \cite{pav_moments_2015,taubock_complex-valued_2012},
and \cite{karseras_caution:_2014}

@@

The question of differentiability of complex-valued networks is tricky, since they usually
use non-differentiable non-linearities. However, a more severe problem stems from the loss
objective being real-valued. Indeed, if $f\colon \cplx \to \cplx$ were $\cplx$-differentiable
(holomorphic), the composition of $f$ with a non-trivial $h \colon \cplx\to \real$ makes it
not $\cplx$-differentiable with respect parameters, since $h$ violates Cauchy-Riemann conditions.
This problem is dealt with by using Wirtinger, or $\cplx\real$ calculus, which subsumes
$\cplx$-calculus, and is applicable to non-holomorphic functions of $\cplx$-argument,
\cite{adali_complex-valued_2011}. As a basis for backpropagation algorithm it was considered
in \cite{benvenuto_complex_1992} and \cite{trabelsi_deep_2017}.

Basically $\cplx\real$ calculus regard $f(z)$ as a function of two independent variables $z$ 
and its conjugate $\conj{z}$ and symbolically defines derivative operators $
  \tfrac{\partial}{\partial z}
$ and $
  \tfrac{\partial}{\partial \conj{z}}
$, that satisfy linearity, product and chain rules, and respect complex conjugation. The important
property is that, if we identify $f(z) = f(u + jv) = F(u, v)$, then the $\cplx\real$ calculus
rules differential of $f$ as a $\cplx$ function coincides its differential as an $\real^2$
function. This makes it possible to use and apply the rules of $\real$-calculus to the networks
represented in paired $\real$ fashion and use existing $\real$-valued backpropagation in autograd
software. The key analytic advantage is that computing derivatives is more streamlined, less
messy, and derivative expressions involving complex variables turn out to be simpler that their
real-imaginary analogues, \cite{adali_complex-valued_2011}.

@@

test on MusicNet, (music transcription [14])

complex high-dimensional time series In-phase Quadrature (IQ)

@@

% TLDR: we use Wirtinger calculus!

% essential intro into Wirtinger calculus (CR)
% https://math.stackexchange.com/a/444493
Wirtinger calculus relies on the natural identification of $\cplx$ with $\real \times \real$.
Let's consider $
  f\colon \cplx \to \cplx
$ and treat it as an equivalent function $\real^2 \to \cplx$ via $
  f(z) = f(u + jv) = F(u, v)
$. Wirtinger calculus defines two formal derivative operators
$
  \tfrac{\partial}{\partial z}
    = \tfrac12 \bigl(
      \tfrac{\partial}{\partial u}
      - j \tfrac{\partial}{\partial v}
    \bigr)
$ and $
  \tfrac{\partial}{\partial \conj{z}}
    = \tfrac12 \bigl(
      \tfrac{\partial}{\partial u}
      + j \tfrac{\partial}{\partial v}
    \bigr)
$ and the differential of $f$ at $z \in \cplx$ as
$$
df(z) \big\vert_{z=u + jv}
  = \frac{\partial f}{\partial z} dz
    + \frac{\partial f}{\partial \conj{z}} d\conj{z}
   % = \frac12\bigl(
   %    \frac{\partial}{\partial u}
   %    - j \frac{\partial}{\partial v}
   % \bigr) F (du - j dv)
   % + \frac12\bigl(
   %    \frac{\partial}{\partial u}
   %    + j \frac{\partial}{\partial v}
   % \bigr) F (du - j dv)
   % = \frac12 \bigl(
   %    \frac{\partial F}{\partial u} du - j \frac{\partial F}{\partial v} du
   %    + j \frac{\partial F}{\partial u} dv + \frac{\partial F}{\partial v} dv
   % \bigr)
   % + \frac12 \bigl(
   %    \frac{\partial F}{\partial u} du + j \frac{\partial F}{\partial v} du
   %    - j \frac{\partial F}{\partial u} dv + \frac{\partial F}{\partial v} dv
   % \bigr)
   = \frac{\partial F}{\partial u} du
     + \frac{\partial F}{\partial v} dv
   = dF(u, v)
  \,, $$
where $dz = du + j dv$ and $d\conj{z} = du - j dv$. This implies that the complex argument
and its conjugate are treated as independent variables. In this notation Cauchy-Riemann
conditions are expressed as $
  \tfrac{\partial f}{\partial \conj{z}} = 0
$, or $
  -j \tfrac{\partial F}{\partial v} = \tfrac{\partial F}{\partial v}
$. These conditions impose a rigid structure on real and imaginary parts of $F(u, v)$, namely $
  \tfrac{\partial F_{\Re }}{\partial u} = \tfrac{\partial F_{\Im }}{\partial v}
$ and $
  \tfrac{\partial F_{\Re }}{\partial v} = - \tfrac{\partial F_{\Im }}{\partial u}
$. Thus Wirtinger calculus subsumes the usual $\cplx$-calculus on holomorphic functions,
when $
  \tfrac{\partial f}{\partial \conj{z}} = 0
$, i.e. $
  f(z)
    % = g(z, \conj{z})
    = F(\tfrac12 (z + \conj{z}), \tfrac1{j 2} (z - \conj{z}))
$ does not depend on $\conj{z}$. The usual rules of calculus, like chain and product rules,
follow directly from the definition of the operators, e.g.
$$
  \frac{\partial (f\circ g)}{\partial z}
    = \frac{\partial f(g(z))}{\partial g} \frac{\partial g(z)}{\partial z}
    + \frac{\partial f(g(z))}{\partial \conj{g}} \frac{\partial \conj{g(z)}}{\partial z}
    % = \nabla G \nabla F
  \,. $$
% Show that $dh(z) = df(g(z)) dg(z)$.
% $h(z) = f(g(z)) = F(G_r(u, v), G_i(u, v)) = H(u, v)$
% $$
% dH
%   = \tfrac{\partial F}{\partial x} \tfrac{\partial G_r}{\partial u} du
%   + \tfrac{\partial F}{\partial x} \tfrac{\partial G_r}{\partial v} dv
%   + \tfrac{\partial F}{\partial y} \tfrac{\partial G_i}{\partial u} du
%   + \tfrac{\partial F}{\partial y} \tfrac{\partial G_i}{\partial v} dv
%   = \tfrac{\partial F}{\partial r} d\Re g(z)
%   + \tfrac{\partial F}{\partial i} d\Im g(z)
%   = df(\Re g(z) + j\Im g(z)) d g(z)
%   = df(g(z)) d g(z)
%   \,. $$

Since in machine learning the target loss is typically an empirical real-valued risk, which
is minimized, and we let's derive the expression for the gradient and make sure that it is
compatible with real-valued calculus upon which current auto-differentiation tools are built.
For a real-valued $f\colon \cplx \to \real$ we have $\conj{f} = f$, whence $
  \tfrac{\partial f}{\partial \conj{z}}
    = \tfrac{\partial \conj{f}}{\partial \conj{z}}
    = \conj{\tfrac{\partial f}{\partial z}}
$. Hence
$$
df
  = \tfrac{\partial f}{\partial z} dz
    + \tfrac{\partial f}{\partial \conj{z}} d\conj{z}
  % = \conj{\tfrac{\partial \conj{f}}{\partial \conj{z}}} dz
  % + \conj{\conj{\tfrac{\partial f}{\partial \conj{z}}} dz}
  = 2 \Re \Bigl(
    \conj{\tfrac{\partial f}{\partial \conj{z}}} dz
  \Bigr)
  = 2 \Re \bigl(
    \tfrac{\partial f}{\partial z} dz
  \bigr)
  \,. $$
Hence the gradient of $f$ at $z$ is given by $
  \tfrac{\partial f}{\partial \conj{z}}
    = \tfrac{\partial F}{\partial u}
      + j \tfrac{\partial F}{\partial v}
$, meaning that we can use gradients with respect ral and imaginary parts independently. Thus
using Wirtinger calculus makes it possible to reuse $\real$-backpropagation for $\cplx$-networks.
